---
title: "Double knockout (DKO, *grin2a* and *grin2b*): NMDA-EPSC peak" 
output:
  html_document: default
  pdf_document: default
---
  
**Document format**  
Great tutorial for [R Markdown] (https://rmarkdown.rstudio.com/articles_intro.html)  
R Markdown reports rely on three frameworks  
* markdown for formatted text  
* knitr for embedded R code  
* YAML for render parameters  
  
**This analysis uses linear mixed modelling with restricted maximum likelihood (REML)**  
  
**Load package requirements**  
```{r, echo=TRUE, message=FALSE, results='hide'}
if (!require(package="tidyverse")) utils::install.packages("tidyverse")
library(tidyverse) 
if (!require(package="lme4")) utils::install.packages("lme4")
library(lme4)  
if (!require(package="HLMdiag")) utils::install.packages("HLMdiag")
library(HLMdiag)
if (!require(package="parameters")) utils::install.packages("parameters")
library(parameters) 
if (!require(package="car")) utils::install.packages("car")
library(car)  
if (!require(package="performance")) utils::install.packages("performance")
library(performance) 
if (!require(package="BayesFactor")) utils::install.packages("BayesFactor")
library(BayesFactor) 
if (!require(package="bayestestR")) utils::install.packages("bayestestR")
library(bayestestR) 
if (!require(package="stats")) utils::install.packages("stats")
library(stats)
if (!require(package="MuMIn")) utils::install.packages("MuMIn")
library(MuMIn)
if (!require(package="pCalibrate")) utils::install.packages("pCalibrate")
library(pCalibrate)
if (!require(package="afex")) utils::install.packages("afex")
library(afex)
if (!require(package="emmeans")) utils::install.packages("emmeans")
library(emmeans)
if (!require(package="multcomp")) utils::install.packages("multcomp")
library(multcomp)
if (!require(package="knitr")) utils::install.packages("knitr")
library(knitr)
if (!require(package="kableExtra")) utils::install.packages("kableExtra")
library(kableExtra)
if (!require(package="ggplot2")) utils::install.packages("ggplot2")
library(ggplot2)
if (!require(package="qqplotr")) utils::install.packages("qqplotr")
library(qqplotr)
if (!require(package="gridExtra")) utils::install.packages("gridExtra")
library(gridExtra)
if (!require(package="ggforce")) utils::install.packages("ggforce")
library(ggforce)
if (!require(package="devEMF")) utils::install.packages("devEMF")
library(devEMF)
if (!require(package="effectsize")) utils::install.packages("effectsize")
library(effectsize)
```

**Read text in from file**
```{r}
Data <- read.delim("../data/n2a_dko_mutant_nmdar.dat", header = TRUE)
# Implicit nesting (required for anovaBF)
Data %>% 
  rename(genotype=mutation) %>% 
  filter(grepl('DKO|DKODQP', genotype)) %>%
  mutate(genotype = as.factor(genotype)) %>% 
  mutate(animal = paste0(as.numeric(genotype),animal)) %>% 
  mutate(slice = paste0(animal,slice)) %>% 
  mutate(pair = paste0(slice,pair)) %>% 
  mutate(pair = factor(pair)) -> Data
```

**Factor encoding**
```{r}  
Data$genotype <- as.factor(Data$genotype)
Data$transfection <- as.factor(Data$transfection)
Data$animal <- as.factor(Data$animal)
Data$slice <- as.factor(Data$slice)
Data$pair <- as.factor(Data$pair)
```

**Set genotype WT and transfection - as reference levels**
```{r}
Data$genotype <- factor(Data$genotype, levels=c("DKO","DKODQP"))
Data$transfection <- factor(Data$transfection, levels=c("-","+"))
```

**lmer settings**
```{r}
settings <- lmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4), boundary.tol=0)
```


**Fit a mixed linear model**
```{r, results='asis'}  

# Initialize
variates <- c("peak")
l <- length(variates)

for (i in 1:l) {
  
variates[i] -> resp  

cat('\n\n\n# Analysis of',resp,'\n\n')

# Plot data
# colours selected from:
#  > library(scales)
#  > show_col(hue_pal()(9))
p1 <- Data %>%
    mutate(genotype_jittered = jitter((as.numeric(genotype)+(as.numeric(transfection)-1)/2.5), 0.5),
           grouping=interaction(pair, genotype)) %>%
    mutate(genotype_transfection = as.numeric(genotype)+(as.numeric(transfection)-1)/2.5) %>%
    ggplot(aes(x=genotype, y=!!sym(resp), group=grouping, color=transfection)) + 
    geom_blank() +
    geom_line(aes(genotype_jittered), alpha=0.33, color="grey") +
    geom_point(aes(genotype_jittered), alpha=0.9, shape = 16) +
    scale_color_manual(values=c("grey","#00BA38")) +
    stat_summary(mapping = aes(x=genotype_transfection,y=!!sym(resp)), fun.data="median_hilow", fun.args = list(conf.int=0.5), geom="linerange", color="black", size=1.0,inherit.aes=FALSE) + 
    stat_summary(mapping = aes(x=genotype_transfection,y=!!sym(resp)), fun="median", geom="point", shape=21, fill="white", color="black", size=2.5, stroke=1, inherit.aes=FALSE) +
    ylab(resp) +
    ggtitle("a") +
    theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1),axis.line = element_line(colour="black"),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_blank(),
          panel.background = element_blank(),
          legend.title = element_blank(),
          legend.position = "top")
p2 <- Data %>% 
    pivot_wider(c(genotype,pair,!!sym(resp)),names_from=transfection,values_from=!!sym(resp)) %>% 
    mutate(ratio = `+`/`-`) %>%
    ggplot(aes(x=genotype, y=ratio, colour=genotype)) +
    geom_sina(alpha=0.9, shape = 16) + 
    scale_color_manual(values=c("#D39200","#93AA00")) +
    stat_summary(fun.data="median_hilow", fun.args = list(conf.int=0.5), geom="linerange", color="black", size=1.0) + 
    stat_summary(fun="median", geom="point", shape=21, fill="white", color="black", size=2.5, stroke=1) +
    ylab("ratio") +
    ggtitle("b") +
    theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1),axis.line = element_line(colour="black"),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_blank(),
          panel.background = element_blank(),
          legend.position = "none")
grid.arrange(p1, p2, nrow = 1, ncol = 2, top=sprintf("Summary plots of the data for: %s\n",resp))

# Fit full model optimizing the Maximum Likelihood criterion
contrasts(Data$genotype) <-  rbind(-1,1)/2 
attr(Data$genotype,"contrasts") %>%
  as.data.frame() %>%
  rownames_to_column(var = "genotype") %>% 
  knitr::kable(caption = sprintf("**Matrix of contrasts on genotype: %s**",resp), digits = 2) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
  print()
contrasts(Data$transfection) <-  rbind(-1,1)/2
attr(Data$transfection,"contrasts") %>% 
  as.data.frame() %>%
  rename(contrast = "V1") %>%
  rownames_to_column(var = "transfection") %>% 
  mutate_at("transfection", str_replace_all, pattern = "\\+", replacement = "\\\\+")  %>% 
  mutate_at("transfection", str_replace_all, pattern = "\\-", replacement = "\\\\-")  %>% 
  knitr::kable(caption = sprintf("**Matrix of contrasts on transfection: %s**",resp), digits = 2) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
  print()  
# Without random effects
formula <- sprintf("log(%s) ~ genotype * transfection", resp)
lm_model <- lm(formula, data = Data, na.action = "na.fail") 
# With random effects
formula <- sprintf("log(%s) ~ genotype * transfection + (1|animal/slice/pair)", resp)
lmer_model <- lmer(formula, data = Data, REML = FALSE, control = settings, na.action = "na.fail") 
# Compare models using information theory
AIC(lm_model,lmer_model) %>% 
  mutate(weight = exp(-0.5*(AIC-min(AIC)))) %>%
  mutate(weight = weight/sum(weight)) %>%
  knitr::kable(caption = sprintf("**Akaike information criterion (AIC) and Akaike weights for the full model with and without the random effects: %s**",resp), digits = 2) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
  print()
BIC(lm_model,lmer_model) %>% 
  mutate(weight = exp(-0.5*(BIC-min(BIC)))) %>%
  mutate(weight = weight/sum(weight)) %>%
  knitr::kable(caption = sprintf("**Bayesian information criterion (BIC) and Schwarz weights for the full model with and without the random effects: %s**",resp), digits = 2) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
  print()
# Compare residual standard
data.frame(sigma = c(sigma(lm_model), sigma(lmer_model)), row.names=c("lm_model","lmer_model"))  %>%
  mutate(percent = 100*sigma/max(sigma)) %>%
  knitr::kable(caption = sprintf("**Residual standard deviation for the full model with and without the random effects: %s**",resp), digits = 2) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
  print()
# Assign lmer_model as model
model <- lmer_model

# Evaluate various other feasible models (with the different fixed effects) using information theory
# Compare model adequacy using the Akaike Information Criterion (AIC) 
# AIC represents the amount of information lost when we use a model to approximate reality.
# The lower the value of AIC, the less the amount of information lost in using that model.
model.eval1 <- dredge(model, rank = "AIC") %>% 
  arrange(desc(df))
model.eval1 %>%
  as.data.frame() %>%
  mutate_if(is.double, formatC, format='f',digits = 2) %>%
  mutate_all(str_replace_all, pattern = "\\+", replacement = "\\\\+")  %>%  
  knitr::kable(caption = sprintf("**Akaike information criterion (AIC) and Akaike weights for the various possible models: %s**",resp), digits = 2) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
  print()

# The Bayesian Information Criterion (BIC) measures the degree of belief that a certain model
# is the true data-generating model. The lower the value of BIC, the greater the belief.
# The BIC puts a unit information prior prior on parameters, which can be thought of as a
# prior distribution that contains the same amount of information as a single, typical 
# observation. This prior is quite diffuse.
model.eval2 <- dredge(model, rank = "BIC") %>% 
  arrange(desc(df))
model.eval2 %>%
  as.data.frame() %>%
  mutate_if(is.double, formatC, format='f',digits = 2) %>%
  mutate_all(str_replace_all, pattern = "\\+", replacement = "\\\\+")  %>%  
  knitr::kable(caption = sprintf("**Bayesian information criterion (BIC) and Schwarz weights for the various possible models: %s**",resp), digits = 2) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
  print()

# Plot results of model evaluation
pd <- position_dodge(0.33)
p1 <- data.frame(cbind(model.eval1$df,model.eval1$weight,model.eval2$weight)) %>% 
    as_tibble() %>%
    rename(df="X1") %>%
    rename(AIC="X2") %>%
    rename(BIC="X3") %>%
    pivot_longer(cols=2:3, names_to = "IC", values_to = "weight") %>%
    ggplot(aes(x = df, y = weight, colour=IC, shape=IC)) +
    geom_line(position = pd) +
    geom_point(position = pd, size=4) +
    ggtitle(sprintf("Model evaluation using Information Theory for: log(%s)", resp)) +
    xlab("Model degrees of freedom") +
    ylab("Weight") +
    theme(legend.title = element_blank())
grid.arrange(p1, nrow = 1, ncol = 1)

# Update model to optimize the Restricted Maximum Likelihood criterion
model <- update(model, REML = TRUE)

# Checking model assumptions
resid = residuals(model)
n = length(resid)
stdev = sqrt((n-1)/n) * sd(resid) # standard deviation with denominator n
std_resid = resid/stdev
p1 <- ggplot(Data, aes(x = fitted(model), y = std_resid)) +
  geom_point() +
  ggtitle("a") +
  xlab("Fitted values") + ylab("Standardized Residuals") +
  geom_hline(yintercept = 0) +
  geom_quantile(formula=y~x, color="#619CFF", size=1) +
  geom_smooth(method="loess", formula = y ~ x, color="#F8766D", size=1, se=FALSE)
p2 <- ggplot(Data, aes(x = std_resid)) +
  geom_histogram(aes(y=..density..), binwidth = 0.9*n^(-1/5), fill="#619CFF", alpha=0.33)  +
  geom_density(kernel="gaussian", alpha=0, color="#619CFF", size=1) +
  ggtitle("b") +
  xlab("Standardized Residuals") + ylab("Density") +
  geom_vline(xintercept = 0) +
  geom_function(fun = dnorm, args = list(mean=0, sd=1), col = "#F8766D", size = 1)
p3 <- ggplot(Data, aes(sample = std_resid)) +
  geom_qq_band(distribution = "norm", bandType = "ts", mapping = aes(fill = "TS"), fill="#619CFF", alpha = 0.33) +
  stat_qq() + 
  stat_qq_line(color="#F8766D",size=1) +
  ggtitle("c") +
  xlab("Normal Quantiles") + ylab("Sample Quantiles") 
infl <- hlm_influence(model, level="pair:(slice:animal)")
p4 <- infl %>% 
  mutate(influential = cooksd > 1.0) %>% 
  ggplot(aes(x=`pair:(slice:animal)`,y=cooksd, color=influential)) + 
  geom_segment(aes(x=`pair:(slice:animal)`, xend=`pair:(slice:animal)`, y=0, yend=cooksd)) + 
  geom_point() + 
  scale_color_manual(values=c("#619CFF","#F8766D")) + 
  ylab("Cook's distance") +
  ggtitle("d") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none",
        panel.background = element_rect(color="#EBEBEB"),
        panel.grid = element_blank(),
        panel.grid.minor.y = element_line(color = "white", size=0.25),
        panel.grid.major.y = element_line(color = "white", size=0.5),
        axis.line = element_blank(),
        axis.line.x = element_line(size = 0.5, colour = "black"))
grid.arrange(p1, p2, p3, p4, nrow=2, ncol=2, top=sprintf("Plots of standardized model residuals and Cook's distances: %s\n",resp))

# Calculate ANOVA table for the fitted model (Type III sum of squares) 
car::Anova(model, type = 3, test.statistic = "F") %>%       # Uses Kenward-Roger degrees of freedom
  as.data.frame() %>%
  rownames_to_column(var="Source") %>%
  filter(Source != "(Intercept)") -> aov

# Calculate Bayes Factors for ANOVA and append them to the ANOVA data frame
# Inclusion Bayes Factor based on matched models (prior odds uniform-equal)
Data %>% 
  mutate(logresp = log(!!sym(resp))) %>%
  as.data.frame() -> Data
set.seed(123456)
anovaBF(logresp ~ genotype * transfection + animal + slice + pair, 
                 whichRandom = c("animal","slice","pair"), 
                 whichModels = "withmain", 
                 iterations = 20000,
                 data = Data) %>%
  bayesfactor_inclusion(match_models = TRUE) %>% 
  as.data.frame() %>% 
  na.omit() %>%    # removes the (nuisance) random factors
  mutate(BF = exp(log_BF)) %>%
  mutate_at("BF", formatC, format='g',digits = 3) %>% 
  dplyr::select(BF) %>% 
  unlist() -> aov$BF

# Display ANOVA table
aov %>% 
  mutate(`Pr(>F)` = afex::round_ps_apa(`Pr(>F)`)) %>%       # format p values as APA style
  knitr::kable(caption = sprintf("**ANOVA table (Type III Wald F tests with Kenward-Roger df) and Bayes factors for fixed effects: %s**",resp), digits = 2) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
  print()

# Calculate intraclass correlation coefficients (ICC) for the random effects
icc(model, by_group=TRUE, tolerance=0) %>% 
  as.data.frame() %>% 
  mutate(N = ngrps(model)) %>%
  rbind(.,c("residual",1-sum(.$ICC),nobs(model))) %>%
  mutate(ICC = as.numeric(ICC)) %>%                               
  knitr::kable(caption = sprintf("**Intraclass correlation coefficients for random effects: %s**",resp), digits = 3) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
  print()
  
# Calculated estimated marginal means, By default, emmeans uses Kenward-Roger's method for estimating the degrees of freedom
emm <- emmeans(model, ~ genotype * transfection, data = Data, tran = 'log', type = 'response')
emm %>% 
  summary(calc = c(n = ".wgt.")) %>%
  as.data.frame() %>%
  mutate_at("transfection", str_replace_all, pattern = "\\+", replacement = "\\\\+")  %>% 
  mutate_at("transfection", str_replace_all, pattern = "\\-", replacement = "\\\\-")  %>% 
  relocate(df, .before = response) %>%
  dplyr::select(-SE) %>%
  knitr::kable(caption = sprintf("**Estimated marginal means with 95%% confidence intervals: %s**",resp), digits = 2) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
  print()

# Calculate overall average for untransfected neurons
emmeans(model, ~ genotype * transfection, data = Data) %>%
  as.data.frame() %>%
  filter(transfection == "-") %>%
  dplyr::select(emmean) %>% 
  colMeans() %>%
  exp() %>%
  sprintf("**Overall average of %s for untransfected neurons**: %.2f",resp,.) %>%
  print()

# Calculate transfected/untransfected ratios
emm.transfection <- contrast(emm, method = "trt.vs.ctrl", interaction = FALSE, by = 'genotype', adjust = "none")
emm.transfection %>%
  confint() %>%
  as.data.frame() %>%
  relocate(df, .before = ratio) %>%
  dplyr::select(-SE) %>%
  knitr::kable(caption = sprintf("**Estimated marginal means with 95%% confidence intervals for transfected/untransfected ratios: %s**",resp), digits = 2) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
  print()

# 95% confidence intervals for interaction contrasts
emm.interaction <- contrast(emm, method = "trt.vs.ctrl", interaction = TRUE, adjust = "none")
emm.interaction %>%
  confint() %>% 
  relocate(df, .before = ratio) %>%
  dplyr::select(-SE) %>%
  knitr::kable(caption = sprintf("**95%% confidence intervals for contrasts: %s**",resp), digits = 2) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
  print()

# Standardized effect sizes (*r*) for interaction contrasts
# Methods used the same as this server: https://easystats4u.shinyapps.io/statistic2effectsize/
emm.interaction %>% 
    as.data.frame() %>% 
    mutate(n = df+nrow(.)+1) %>% 
    mutate(r = t_to_r(t.ratio, df)$r) %>% 
    mutate(z = atanh(r),
           SE = 1/sqrt(n-3),
           CI = sprintf("[%.2f, %.2f]",
                        LL = tanh(z - 1.96*SE),
                        UL = tanh(z + 1.96*SE))) %>%
    dplyr::select(-c(ratio,SE,df,null,t.ratio,p.value,z)) %>%
    knitr::kable(col.names = c("genotype",
                               "transfection",
                               "*n*",
                               "*r*",
                               "95% *CI*"),
                 caption = sprintf("**Standardized effect sizes (*r*) for contrasts: %s**",resp), digits = 2) %>% 
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
    print()

posthoc = FALSE

if (posthoc == TRUE) {
  
# p-values and maximum Bayes Factors for interaction contrasts 
# Dunnett's step-down adjustment to control FWER on p-values (using multcomp package)
# Chapter 4.1.2 in Bretz, F., Hothorn, T. and Westfall, P. (2011) Multiple Comparisons Using R. Taylor and Frances Group, LLC.
emm.interaction %>%   
    as.glht() %>%
    summary(test = adjusted(type = "free")) -> glht.out  
emm.interaction %>%   
  as.data.frame() %>%
  dplyr::select(-SE) %>%
  mutate(p.adj = glht.out$test$pvalues) %>%
  mutate(p.adj = sapply(p.adj,max,.Machine$double.eps)) %>%
  mutate(maxBF = 1/pCalibrate(p.adj,"exploratory")) %>%
  mutate_at("maxBF", formatC, format='g',digits = 3) %>%
  mutate(p.value = afex::round_ps_apa(p.value)) %>%
  mutate(p.adj = afex::round_ps_apa(p.adj)) %>%
  knitr::kable(caption = sprintf("**Hypothesis testing on interaction parameters (Dunnett's step-down p-value adjustment): %s**",resp), digits = 2) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
  print()

}

# Replot data with 95% confidence intervals
emf(sprintf("../img/%s_%s.emf","n2a_dko_nmdar",resp), width=2.7, height=3.5)
emm %>%
    as.data.frame() %>%
    mutate(genotype_transfection = as.numeric(genotype)+(as.numeric(transfection)-1)/2.5) -> emm_df
p1 <- Data %>%
    mutate(genotype_jittered = jitter((as.numeric(genotype)+(as.numeric(transfection)-1)/2.5), 0.5),
           grouping=interaction(pair, genotype)) %>%
    mutate(genotype_transfection = as.numeric(genotype)+(as.numeric(transfection)-1)/2.5) %>%
    ggplot(aes(x=genotype, y=!!sym(resp), group=grouping, color=transfection)) + 
    geom_blank() +
    geom_line(aes(genotype_jittered), alpha=0.3, color="grey", size=0.75) +
    geom_point(aes(genotype_jittered), alpha=0.6, shape = 16, size=1.25) +
    scale_color_manual(values=c("grey","#00BA38")) +
    scale_fill_manual(values=c("grey","#00BA38")) +
    geom_crossbar(data = emm_df, 
                    aes(x=genotype_transfection, y=response, ymin=`lower.CL`, ymax=`upper.CL`, fill=transfection), 
                    color="black", alpha=0.5, size=0.5, fatten=1, width=0.3, inherit.aes=FALSE) + 
    ylab(resp) +
    theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1),axis.line = element_line(colour="black"),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_blank(),
          panel.background = element_blank(),
          legend.title = element_blank(),
          legend.position = c(0.5, 1.06),
          legend.direction = "horizontal",
          text = element_text(size=14))
emm.transfection %>%
    confint() %>%
    as.data.frame() -> emm.transfection_df
p2 <- Data %>% 
    pivot_wider(c(genotype,pair,!!sym(resp)),names_from=transfection,values_from=!!sym(resp)) %>% 
    mutate(ratio = `+`/`-`) %>%
    ggplot(aes(x=genotype, y=ratio, colour=genotype)) +
    geom_sina(alpha=0.6, shape=16, size=1.25, maxwidth=0.5) + 
    geom_crossbar(data = emm.transfection_df, 
                           aes(x=genotype, y=ratio, ymin=`lower.CL`, ymax=`upper.CL`, fill=genotype), 
                    color="black", alpha=0.5, size=0.5, fatten=1, width=0.8, inherit.aes=FALSE) +
    scale_color_manual(values=c("#D39200","#93AA00")) +
    scale_fill_manual(values=c("#D39200","#93AA00")) +
    ylab("ratio") +
    ylim(0,1.4) + 
    theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1), axis.line = element_line(colour="black"),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_blank(),
          panel.background = element_blank(),
          legend.position = "none",
          text=element_text(size=14))
grid.arrange(p1, p2, layout_matrix=rbind(c(1,2)), top=sprintf("Summary plots of the data with 95%% confidence intervals: %s\n",resp))
dev.off() #turn off device and finalize file

}

```


